---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(corrplot)
library(MASS)
library(car)
```



```{r}
le = read.csv('Life Expectancy Data.csv')
```

# Data Dictionary  
Adult Mortality (Numeric) - # mortalities of adults(15-60) out of 1000 total mortalities.(legal values 0-1000).  

infant deaths (Numeric) - # mortalities of infants(0-1) out of 1000.(legal values 0-1000).  

Alcohol (Numeric) - avg consumption of pure alcohol. (age>15), (leagal values >=0).  

percentage expenditure (Numeric) - how much spend on health as part of GDP.(legal values <= GDP of the same key).  

Hepatitis B (Ratio %) - number of 1 year olds with Hepatitis B immunization over all 1 year olds in population in percantage.  

Measles (Numeric) - number of reported Measles cases per 1000 population, numbers indicate per population(much more than 1000)  

BMI (Numeric) - average Body Mass Index.  

under-five deaths (Numeric) - number of people under 5 year's old that died out of 1000.  

Polio (Ratio %) - number of 1 year olds immunized from Polio of out all 1 year olds.  

Total expenditure (Ratio %) - General government expenditure on health as a percentage of total government expenditure.  

Diphtheria (Ratio %) - Diphtheria tetanus toxoid and pertussis (DTP3) immunization rate of 1 year olds in percantage.  

HIV/AIDS (Numeric) - deaths per 1000 live births caused by HIV/AIDS for people ages 0-4.  

GDP (Numeric) - Gross Domestic Product per person($).  

Population (Numeric) - how many people lived in a country in a specific year.  

thinness 1-19 years (Ratio %) - thinness ratio among children and adolescents for Age 10 to 19, name should be fix.

thinness 5-9 yearss (Ratio %) - thinness ratio among children and adolescents for Age 5 to 9.

Income composition of resources (Ratio %) - Human Development Index in terms of income composition of resources (index ranging from 0 to 1)

schooling (Numeric) - average number of schooling years.  

# Initial EDA

```{r}
head(le)
```

```{r}
summary(le)
```

**There are values out of range (infant mortality 1800, max is 1000) multiple NA values**

### Visual EDA - Histogram
Let's see how our data is distributed

```{r}
le %>%
  keep(is.numeric) %>%
  gather() %>%
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = 'free') +
  geom_histogram()
```

A lot of values are very Skewed. 

There is a categorical variable for devloped or developing nations, let's look at the difference in life expectancy in the two groups.  

```{r}
le %>%
  group_by(Status) %>%
  summarise(Life_Exp_avg = mean(Life.expectancy, na.rm = TRUE))
```

More than a 10 percent difference between groups.  

Let's look at numerical correlations.  

```{r}
le %>%
  keep(is.numeric) -> le.num
round(cor(as.data.frame(na.omit(le.num))),2) -> C
png(file = 'corr.png', res = 300, width = 4500, height = 4500)
corrplot(C, method='circle', type = 'lower')
```


```{r}
cor(as.data.frame(na.omit(le.num)))[, 2:3]
```

## Clean Data

**replace NA values with column average**

```{r}
for(i in 4:ncol(le)){
  le[is.na(le[,i]), i] <-mean(le[,i], na.rm = TRUE)
}
```
check results
```{r}
summary(le)
```

**use reason to drop columns**
+ Country name has no bearing on life expectancy, drop it  
+ percentage.expenditure is too highly correlated with GDP, let's wee which one to keep:  

```{r}
ggplot(le, aes(x = GDP)) + geom_histogram()
ggplot(le, aes(x=percentage.expenditure)) + geom_histogram() 
ggplot(le, aes(x = GDP)) + geom_boxplot()
ggplot(le, aes(x = percentage.expenditure)) + geom_boxplot()
```



GDP seems slightly better distributed, drop percentage.expenditure  
+ under.five.deaths and infant.deaths are too highly correlated, drop infant.deaths, because all the data in under.five is included in infant.deaths  
+ thinness.5.9.years data in included in thinness..1.19.years, drop 5.9  
+ Adult Mortality is direct measure of target variable, drop it  
+ Income.composition.of.resources is too highly correlated with income.composition

### Drop Columns 
```{r}
le %>%
  dplyr::select(-c(Adult.Mortality, infant.deaths, percentage.expenditure, thinness.5.9.years, Income.composition.of.resources)) -> le.drop

```

**check for illegal values  **
```{r}
# check for under 5 deaths greater than 1000
le.drop %>%
  filter(under.five.deaths <= 1000) -> le.clean
```

the following columns are supposed to be ratios:
```{r}
ratios <- c(le.clean$Hepatitis.B, le.clean$Polio, le.clean$Total.expenditure, le.clean$Diptheria, le.clean$thinness..1.19.years, le.clean$Income.composition.of.resources )
le.clean %>%
  dplyr::mutate(Hepatitis.B = round(Hepatitis.B/100, 3)) %>%
  dplyr::mutate(Polio = round(Polio/100, 3)) %>%
  dplyr::mutate(Total.expenditure =round((Total.expenditure/10),3)) %>%
  dplyr::mutate(Diphtheria = round(Diphtheria/100,3)) %>%
  dplyr::mutate(Thinness = round(thinness..1.19.years/100, 3)) %>%
  dplyr::select(-thinness..1.19.years)  -> le.clean

```



Let's split data into train and test
```{r}
set.seed(7)
tt <- sample(1:2, size = nrow(le.clean), replace = TRUE, prob=c(0.9, 0.1))
le.train <- le.clean[tt == 1,]
le.test <- le.clean[tt == 2,]
```

Let's train a model using step wise by AIC
```{r}
le.train.emptyMod <- lm(Life.expectancy ~ 1, data = le.train[,-c(1,2)])
le.train.fullMod <- lm(Life.expectancy ~ ., data = le.train[,-c(1,2)])
scope = list(lower = formula(le.train.emptyMod), upper = formula(le.train.fullMod))
le.train.FAIC <- step(le.train.emptyMod, scope, direction = 'forward', k=2)
```

**let's look at the model**
```{r}
summary(le.train.FAIC)
```
```{r}
vif(le.train.FAIC)
```


```{r}
png(file = '2.png')
plot(le.train.FAIC, which =2)
```

**lets do some prediction with our test DF**
```{r}
le.predictCon <- predict(le.train.FAIC, le.test, interval = 'confidence')
le.predictPred <-predict(le.train.FAIC, le.test, interval = 'prediction')


```
**concat upper and lower confidence and prediction value to test dataframe**
```{r}
cbind(le.test, 'prediction' = le.predictCon[,1], 
      'confidence.lower' = le.predictCon[,2],
      'confidence.upper' = le.predictCon[,3],
      'prediction.lower' = le.predictPred[,2],
      'prediction.upper' = le.predictPred[,3]) -> predictedValues

```
```{r}
library(ggrepel)
png(file = 'ObsPred.png')
g <- ggplot(predictedValues, aes(y = Life.expectancy, x = prediction))
g + geom_point(size = 1.05) + geom_smooth(method = "lm", se = FALSE, aes(color = 'Regression Line')) +
  geom_abline(intercept = 1.0644, slope = .9923, color = 'darkgreen', se = FALSE, aes(color = 'Confidence Band')) + 
  geom_abline(intercept = -.9698, slope = 1.0063, color = 'darkgreen', se = FALSE,  aes(color = 'Confidence  Band')) +
  geom_abline(intercept = 9.0986, slope = .994, color = 'red', se = FALSE, aes(color = 'Prediction Band')) +
  geom_abline(intercept = -9.109, slope = 1.001, color = 'red', se = FALSE, aes(color = 'Prediction Band')) + labs(
    x = 'Predicted Life Expectancy', y = 'Observed Life Expectancy') +  
    ggtitle('Observed Vs Predicted Life Expectancy') +
    scale_color_manual(name='Regression Model',
                     breaks=c('Regression Line', 'Confidence Band', 'Prediction Band'),
                     values=c('Prediction Band'='red', 'Regression Line'='blue', 'Confidence Band'='darkgreen'))
  
```
```{r}
png(file = 'actual.png')
predictedValues %>%
  dplyr::select(c(Life.expectancy, prediction)) %>%
  ggplot(., aes(y = Life.expectancy)) + geom_boxplot() + ggtitle('Distribution of Actual Life Expectancies') + labs(y = 'Life Expectancy') + scale_y_continuous(limits = c(50,90))
```

```{r}
png(file = 'predict.png')
predictedValues %>%
  dplyr::select(c(Life.expectancy, prediction)) %>%
  ggplot(., aes(y = prediction)) + geom_boxplot() + ggtitle('Distribution of Predicted Life Expectancies') + labs(y = 'Life Expectancy') +  scale_y_continuous(position = 'right', limits = c(50,90))
```




```{r}
t.test(le.test$Life.expectancy, le.predictCon[,1])
```
```{r}
library(ggpubr)
library(rstatix)
library(gginference)
library(MASS)
png(file = 'ttest.png')
ggttest(t.test(le.test$Life.expectancy, le.predictCon[,1]))
```

Model looks good  
**There are 2 variables that can be tailored to an individual can we input numbers for a person and estimate their life expectancy**
+ BMI  
+ Schooling  

**let's change the BMI by 10% on the test data and rerun the prediction and see how the values change**

```{r}
le.test %>%
  mutate(BMI = BMI *1.1) -> le.test.bmi
le.predictBMI <-predict(le.train.FAIC, le.test.bmi, interval = 'prediction')
cbind(le.test, 'Nudge.BMI' = le.test.bmi$BMI, 'prediction.original' = predictedValues$prediction, 'prediction.bmi' = le.predictBMI[,1]) -> nudgeBMI
nudgeBMI

```

```{r}
#calculate average difference
nudgeBMI %>%
  summarise(per_change = mean(abs(prediction.bmi - prediction.original)/prediction.original) * 100)
```

```{r}
g <- ggplot(nudgeBMI, aes(y = prediction.original, x = BMI))
g + geom_smooth(se = FALSE, aes(color = 'Original')) + geom_smooth(aes(x = Nudge.BMI, y = prediction.bmi, color = 'BMI +10%'), se = FALSE) + labs(
  x = 'BMI', y = 'Predicted Life Expectancy') + ggtitle('BMI vs Life Expectancy') +
  scale_color_manual(name='BMI',
                     breaks=c('Original', 'BMI +10%'),
                     values=c('Original'='blue', 'BMI +10%'='red'))
```
```{r}
le.test %>%
  mutate(Schooling = Schooling *1.1) -> le.test.schooling
le.predictSchooling <-predict(le.train.FAIC, le.test.schooling, interval = 'prediction')
cbind(le.test, 'Nudge.Schooling' = le.test.schooling$Schooling, 'prediction.original' = predictedValues$prediction, 'prediction.schooling' = le.predictSchooling[,1]) -> nudgeSchooling
nudgeSchooling
```
```{r}
#calculate average difference
nudgeSchooling %>%
  summarise(per_change = mean(abs(prediction.schooling - prediction.original)/prediction.original) * 100)
```
```{r}
g <- ggplot(nudgeSchooling, aes(y = prediction.original, x = Schooling))
g + geom_smooth(se = FALSE, aes(color = 'Original')) + geom_smooth(aes(x = Nudge.Schooling, y = prediction.schooling, color = 'Schooling +10%'), se = FALSE) + labs(
  x = 'Schooling', y = 'Predicted Life Expectancy') + ggtitle('Schooling vs Life Expectancy') +
  scale_color_manual(name='BMI',
                     breaks=c('Original', 'Schooling +10%'),
                     values=c('Original'='blue', 'Schooling +10%'='red'))
```

**write a function that takes in country schooling and BMI and outputs a life expectancy with 95% confidence levels**

```{r}
expectLife <- function(country,BMI,school){
  
# need to pull status from df
first(le.clean[le.clean$Country == country,]$Status) -> developing
  
  le.clean %>%
    dplyr::select(Country, HIV.AIDS, Diphtheria, under.five.deaths, GDP, Polio, Thinness, Hepatitis.B, Population) %>%
    dplyr::filter(Country == country) %>%
    dplyr::select(-Country) %>%
    summarise_all(mean) %>%
    cbind(., 'Status' = developing, 'BMI' = BMI, 'Schooling' = school) -> newwdata
  prediction <- predict(le.train.FAIC, newwdata, interval = 'confidence')
  prediction[,1]
  
}
expectLife('United States of America', 30, 16)
```

*
**let's see how our predictions change when a single variable changes**

```{r}
# make a df with only modeled and target variable
le.train %>%
  dplyr::select(-Measles, Total.expenditure, -Country,-Year) -> modeledDF
tempDF <- modeledDF
graphDF <- data.frame()
for (i in 3:ncol(modeledDF)){
  
  oldColName <- colnames(modeledDF)[i]
  graphname <- paste0(oldColName, '_Graph')
  for (j in c(.5, .7, .9, 1, 1.1, 1.5, 2, 3)){
    newValues <- modeledDF[oldColName] * j
    tempDF[oldColName] <- modeledDF[oldColName] * j
    predictValues <- predict(le.train.FAIC, modeledDF, interval = 'prediction')
    jFrame <- data.frame(as.factor(rep(j, nrow(modeledDF))),
                         tempDF[oldColName],
                         predictValues[,1])
    graphDF <- rbind(graphDF, jFrame)
  }
 colnames(graphDF) <-  c('Nudge', oldColName, 'Predicted')
 g <- ggplot(graphDF, aes(x = graphDF[,2], y = graphDF[,3], colour = graphDF[,1]))
 g <- g + geom_smooth(se = FALSE) + ggtitle(paste0(oldColName, ' vs. Life Expectancy'))
 g <- g + labs(x = oldColName, y = 'Life Expectancy') + coord_cartesian(xlim =  c(quantile(modeledDF[,oldColName], .25), quantile(modeledDF[,oldColName], .75)))
 
 show(g)
 graphDF <- data.frame()
 tempDF <- modeledDF
 
}
```




# Creating content for presentation  

```{r}
le %>%
  rename('Adult Mortality' = Adult.Mortality,
         'Life Expectancy' = Life.expectancy,
         'Infant Deaths' = infant.deaths,
         'Alcohol Per Week' = Alcohol,
         'Healthcare Spending of GDP' = percentage.expenditure,
         'Hepatitis B Vax Rate' = Hepatitis.B,
         'Measles Cases' = Measles,
         'Todler Mortality' = under.five.deaths,
         'Polio Vax Rate' = Polio,
         'Health Spending % of Budget' = Total.expenditure,
         'DTP3 Vax Rate' = Diphtheria,
         'Infant HIV/AIDS Deaths' = HIV.AIDS,
         'Adolescent Underweight' = thinness..1.19.years,
         'Toddler Underweight' = thinness.5.9.years,
         'Human Developement Index' = Income.composition.of.resources,
         'Education (Years)' = Schooling) %>%
  dplyr::select(-c(Country, Status)) -> D
round(cor(as.data.frame(na.omit(D))),2) -> C
png(file = 'CompleteCorrelation.png')
corrplot(C, method='circle', type = 'lower', color = 'black')

```

le %>%
  keep(is.numeric) -> le.num
round(cor(as.data.frame(na.omit(le.num))),2) -> C
png(file = 'corr.png', res = 300, width = 4500, height = 4500)
corrplot(C, method='circle', type = 'lower')

```{r}
png(file = 'GDPPerPerson.png', res = 500, width = 4500, height = 4500)
ggplot(le, aes(x =(GDP - mean(GDP))/sd(GDP)+ .6)) + geom_histogram(bins = 100)  + ggtitle('Distribution of Healthcare Spending Per Person') + labs(x = 'Percentage of Budget Spent on Healthcare', y = 'Count') + theme(axis.title.x = element_text(size = 15),                                                                                                                                              axis.title.y = element_text(size = 14),
      plot.title = element_text(size = 20))
                                                                                                                                                                            
```



```{r}
png(file = 'HCPerCentExpenditure.png', res = 500, width = 4500, height = 4500)
ggplot(le, aes(x = Total.expenditure * 1000)) + geom_histogram(bins = 100) + ggtitle('Distribution of GDP') + labs(x = 'GDP Per Person ($)', y = 'Count')+
   theme(axis.title.x = element_text(size = 15),                                                                                                                                              axis.title.y = element_text(size = 14),
      plot.title = element_text(size = 20))
```



```{r}
normalized <- function(y){
  x <- (y - min(y))/(max(y) - min(y))
  return(x)
}
le.clean %>%
  mutate(AlcoholScaled = 1 - normalized(Alcohol)) %>%
  mutate(LEScaled = normalized(Life.expectancy)) %>%
  mutate(HepScaled = normalized(Hepatitis.B)) %>%
  mutate(MeaslesScaled = normalized(Measles))%>%
  mutate(BMIScaled = 1 - normalized(BMI))%>%
  mutate(PolioScaled = normalized(Polio))%>%
  mutate(TotalExpScaled = 1-normalized(Total.expenditure))%>%
  mutate(DipScaled = normalized(Diphtheria)) %>%
  mutate(HealthIndex = AlcoholScaled * LEScaled * HepScaled * MeaslesScaled * BMIScaled * PolioScaled * TotalExpScaled * DipScaled) %>%
  dplyr::select(Country, Year, HealthIndex) -> HealthIndexDF
  
  
  
```
```{r}
png(file = 'HealthindexHist.png')
HealthIndexDF %>%
  ggplot(., aes(x = HealthIndex)) + geom_histogram(bins = 500) + xlim(0, .01) + ylim(0,300) +
  ggtitle('Distribution of HealthIndex by Country' )
```


```{r}
HealthIndexDF %>%
  group_by(Country) %>%
  summarise(meanHI = mean(HealthIndex)) %>%
  arrange(desc(meanHI)) %>%
  top_n(5)%>%
  dplyr::select(Country) -> topCountries
png(file = 'healthindexbyyear.png', height = 400, width = 900)
HealthIndexDF%>%
  filter(Country %in% topCountries$Country) %>%
  ggplot(., aes(x = Year, y = HealthIndex, color = Country)) + geom_line() + 
  ggtitle('Health Index Over Time by Country')
```





