---
title: "R Project Notebook"
date: "`r Sys.Date()`"
author: "Justin Meisenhelter"
output: 
  rmdformats::downcute:
    highlight: tango
    df_print: paged
    collapsed: false
---
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(corrplot)
```



```{r}
le = read.csv('Life Expectancy Data.csv')
```

# Data Dictionary  
Adult Mortality (Numeric) - # mortalities of adults(15-60) out of 1000 total mortalities.(legal values 0-1000).  

infant deaths (Numeric) - # mortalities of infants(0-1) out of 1000.(legal values 0-1000).  

Alcohol (Numeric) - avg consumption of pure alcohol. (age>15), (leagal values >=0).  

percentage expenditure (Numeric) - how much spend on health as part of GDP.(legal values <= GDP of the same key).  

Hepatitis B (Ratio %) - number of 1 year olds with Hepatitis B immunization over all 1 year olds in population in percantage.  

Measles (Numeric) - number of reported Measles cases per 1000 population, numbers indicate per population(much more than 1000)  

BMI (Numeric) - average Body Mass Index.  

under-five deaths (Numeric) - number of people under 5 year's old that died out of 1000.  

Polio (Ratio %) - number of 1 year olds immunized from Polio of out all 1 year olds.  

Total expenditure (Ratio %) - General government expenditure on health as a percentage of total government expenditure.  

Diphtheria (Ratio %) - Diphtheria tetanus toxoid and pertussis (DTP3) immunization rate of 1 year olds in percantage.  

HIV/AIDS (Numeric) - deaths per 1000 live births caused by HIV/AIDS for people ages 0-4.  

GDP (Numeric) - Gross Domestic Product per person($).  

Population (Numeric) - how many people lived in a country in a specific year.  

thinness 1-19 years (Ratio %) - thinness ratio among children and adolescents for Age 10 to 19, name should be fix.

thinness 5-9 yearss (Ratio %) - thinness ratio among children and adolescents for Age 5 to 9.

Income composition of resources (Ratio %) - Human Development Index in terms of income composition of resources (index ranging from 0 to 1)

schooling (Numeric) - average number of schooling years.  

# Initial EDA

```{r}
head(le)
```

```{r}
summary(le)
```

**There are values out of range (infant mortality 1800, max is 1000) multiple NA values**

### Visual EDA - Histogram
Let's see how our data is distributed

```{r}
le %>%
  keep(is.numeric) %>%
  gather() %>%
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = 'free') +
  geom_histogram()
```

A lot of values are very Skewed. 

There is a categorical variable for devloped or developing nations, let's look at the difference in life expectancy in the two groups.  

```{r}
le %>%
  group_by(Status) %>%
  summarise(Life_Exp_avg = mean(Life.expectancy, na.rm = TRUE))
```

More than a 10 percent difference between groups.  

Let's look at numerical correlations.  

```{r}
le %>%
  keep(is.numeric) -> le.num
round(cor(as.data.frame(na.omit(le.num))),2) -> C
png(file = 'corr.png', res = 300, width = 4500, height = 4500)
corrplot(C, method='circle', type = 'lower')
```


```{r}
cor(as.data.frame(na.omit(le.num)))[, 2:3]
```

## Clean Data

**replace NA values with column average**

```{r}
for(i in 4:ncol(le)){
  le[is.na(le[,i]), i] <-mean(le[,i], na.rm = TRUE)
}
```
check results
```{r}
summary(le)
```

**use reason to drop columns**
+ Country name has no bearing on life expectancy, drop it  
+ percentage.expenditure is too highly correlated with GDP, let's wee which one to keep:  

```{r}
ggplot(le, aes(x = GDP)) + geom_histogram()
ggplot(le, aes(x=percentage.expenditure)) + geom_histogram() 
ggplot(le, aes(x = GDP)) + geom_boxplot()
ggplot(le, aes(x = percentage.expenditure)) + geom_boxplot()
```



GDP seems slightly better distributed, drop percentage.expenditure  
+ under.five.deaths and infant.deaths are too highly correlated, drop infant.deaths, because all the data in under.five is included in infant.deaths  
+ thinness.5.9.years data in included in thinness..1.19.years, drop 5.9  
+ Adult Mortality is direct measure of target variable, drop it  

### Drop Columns 
```{r}
le %>%
  dplyr::select(-c(Adult.Mortality, infant.deaths, percentage.expenditure, thinness.5.9.years)) -> le.drop

```

**check for illegal values  **
```{r}
# check for under 5 deaths greater than 1000
le.drop %>%
  filter(under.five.deaths <= 1000) -> le.clean
```

the following columns are supposed to be ratios:
```{r}
ratios <- c(le.clean$Hepatitis.B, le.clean$Polio, le.clean$Total.expenditure, le.clean$Diptheria, le.clean$thinness..1.19.years, le.clean$Income.composition.of.resources )
le.clean %>%
  dplyr::mutate(Hepatitis.B = round(Hepatitis.B/100, 3)) %>%
  dplyr::mutate(Polio = round(Polio/100, 3)) %>%
  dplyr::mutate(Total.expenditure =round((Total.expenditure/10),3)) %>%
  dplyr::mutate(Diphtheria = round(Diphtheria/100,3)) %>%
  dplyr::mutate(Thinness = round(thinness..1.19.years/100, 3)) %>%
  dplyr::select(-thinness..1.19.years) %>%
  dplyr::mutate(Income.composition = round(Income.composition.of.resources,3)) -> le.clean

```

Group by country and look for values in fields with more than 10% change over all years  
```{r}
le.clean %>%
  dplyr::select(-Status) %>%
  dplyr::group_by(Country) %>%
  dplyr::summarise(across(where(is.numeric),
                          ~ (max(.x, na.rm = TRUE) - min(.x, na.rm = TRUE))/max(.x, na.rm = TRUE))) -> le.diff  
le.diff
```

```{r}
for(i in 2:ncol(le.diff)){
  le.diff[is.na(le.diff[,i]), i] <- mean(le.diff[,i], na.rm = TRUE)
}
le.diff
```
Lets graph all variables as a histogram and see if any have extreme amounts of large change
```{r}
le.diff %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()

```


Variables where very large percentage changes are suspicious:  
+ Alcohol  
+ BMI  
+ GDP  
+ Income.composition  
+ Income.composition.of.resources
+ Population  
+ under.five.deaths

### Alcohol Variability  

```{r}
le.diff %>%
  filter(Alcohol > .9) %>%
  dplyr::select(Country) -> le.high.alcohol
le.clean %>%
  inner_join(., le.high.alcohol, by = 'Country') 
  
```



**Let's save further datacleanup for if we can't get a good model**
create basic saturated linear model
```{r}
le.saturated = lm(Life.expectancy ~ . - Country, data = le.clean)
summary(le.saturated)

```

**Assessing the assumptions of the model** 
```{r}
plot(le.saturated)
```


**Check VIF of all the variables**  

```{r}
vif(le.saturated)
```
Income composition variables seem to be the highest, but below threshold of 5  

**perform step-wise feature selection**
```{r}
library(MASS)
le.empty = lm(Life.expectancy ~1, data = le.clean)
le.full = lm(Life.expectancy ~ . -Country, data = le.clean)
scope = list(lower = formula(le.empty), upper = formula(le.full))
forwardAIC = step(le.empty, scope, direction= 'forward', k=2)
```

```{r}
summary(forwardAIC)
```
pretty good initial fit, lets try some predictions


### Test and Train

**split dataframe into 90% train and 10% test**
```{r}
set.seed(7)
tt <- sample(1:2, size = nrow(le.clean), replace = TRUE, prob=c(0.9, 0.1))
le.train <- le.clean[tt == 1,]
le.test <- le.clean[tt == 2,]
```

**create DF without country column**
```{r}
le.clean %>%
  dplyr::select(-Country) -> le.cleanNC
```


**do forward AIC stepwise to get model for this data**

```{r}

le.train.emptyMod <- lm(Life.expectancy ~ 1, data = le.cleanNC)
le.train.fullMod <- lm(Life.expectancy ~ ., data = le.cleanNC)
scope = list(lower = formula(le.train.emptyMod), upper = formula(le.train.fullMod))
le.train.FAIC <- step(le.train.emptyMod, scope, direction = 'forward', k=2)
```
**lets compare model from full dataset to model from train dataset**
```{r}
anova(le.train.FAIC, forwardAIC)
```


**The models are identical**

**lets do some prediction with our test DF**
```{r}
le.predictCon <- predict(le.train.FAIC, le.test, interval = 'confidence')
le.predictPred <-predict(le.train.FAIC, le.test, interval = 'prediction')
lm(le.test$Life.expectancy ~ le.predictCon[,1])
```
**lets try and visualize this**
```{r}
plot(le.test$Life.expectancy, le.predictCon[,1], xlab = 'Actual Life Expectancy',
     ylab = 'Prediction', main = 'Predicted versus Actual Life expectancy', col = 'blue')
abline(-3.254, 1.049)
plot(le.test$Life.expectancy, le.predictPred[,1], xlab = 'Actual Life Expectancy',
     ylab = 'Prediction', main = 'Predicted versus Actual Life expectancy', col = 'red')

```
...not great was hoping to see a straight line


**lets two a T-Test to determine how far off our model is**
```{r}
t.test(le.clean$Life.expectancy, le.predictPred[,1])
```
not as low as I was hoping, I guess we have more work to do

## Box Cox transformation  

**let's see which variables could be a candidate for transformation**
```{r}
le.train %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()
```


```{r}
bc = boxCox(le.train.FAIC)
lambda = bc$x[which(bc$y == max(bc$y))]
# transforma dataset

```

```{r}
bctransform <- function(x){
  x^2-1/2
}
le.bc <-le.train
le.bc[5:19] <- lapply(le.bc[5:19],bctransform)
```
**look at transformed values**
```{r}
le.bc %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()
```



## Retrain the model  
```{r}
le.bc.emptyMod <- lm(Life.expectancy ~ 1, data = le.bc[,-c(1,2)])
le.bc.fullMod <- lm(Life.expectancy ~ ., data = le.bc[,-c(1,2)])
scope = list(lower = formula(le.bc.emptyMod), upper = formula(le.bc.fullMod))
le.bc.FAIC <- step(le.bc.emptyMod, scope, direction = 'forward', k=2)
```

**let's see what we got**
```{r}
plot(le.bc.FAIC)
summary(le.bc.FAIC)
```

this is worse, let's forget box cox.
```{r}
le.clean %>%
  keep(is.numeric) -> le.cleanNum
round(cor(as.data.frame(na.omit(le.cleanNum))),2) -> C
png(file = 'corr.png', res = 300, width = 4500, height = 4500)
corrplot(C, method='circle', type = 'lower')
```

income composition is correlated with a lot of variables: income.composition.of.resources., and schooling..let's drop those two and try again  

```{r}
set.seed(7)
le.clean %>%
  dplyr::select(-c(Schooling, -Income.composition.of.resources)) -> le.reduced
tt <- sample(1:2, size = nrow(le.reduced), replace = TRUE, prob=c(0.9, 0.1))
le.train <- le.reduced[tt == 1,]
le.test <- le.reduced[tt == 2,]
```
use forward AIC to train a new model with less variables

```{r}
le.train.emptyMod <- lm(Life.expectancy ~ 1, data = le.train[,-c(1,2)])
le.train.fullMod <- lm(Life.expectancy ~ ., data = le.train[,-c(1,2)])
scope = list(lower = formula(le.train.emptyMod), upper = formula(le.train.fullMod))
le.reduced.FAIC <- step(le.train.emptyMod, scope, direction = 'forward', k=2)
```

lets look at the model

```{r}
summary(le.reduced.FAIC)
```



**lets do some prediction with our test DF**
```{r}
le.predictCon <- predict(le.reduced.FAIC, le.test, interval = 'confidence')
le.predictPred <-predict(le.reduced.FAIC, le.test, interval = 'prediction')
lm(le.test$Life.expectancy ~ le.predictCon[,1])
```
**lets try and visualize this**
```{r}
plot(le.test$Life.expectancy, le.predictCon[,1], xlab = 'Actual Life Expectancy',
     ylab = 'Prediction', main = 'Predicted versus Actual Life expectancy', col = 'blue')
abline(-3.254, 1.049)
plot(le.test$Life.expectancy, le.predictPred[,1], xlab = 'Actual Life Expectancy',
     ylab = 'Prediction', main = 'Predicted versus Actual Life expectancy', col = 'red')

```


```{r}
t.test(le.test$Life.expectancy, le.predictCon[,1])
```




```{r}
options(scipen = 999)
meanround <- function(x){
    round(mean(x),2)

}
  df %>%
    dplyr::filter(Country == 'Germany') %>%
    dplyr::select(5:18) %>%
    summarise_all(meanround) %>%
    rename('Life Expectancy (Years)' = Life.expectancy,
           'Average Drinks Per Week' = Alcohol,
           'Percentage Vax Hepatitis B'=  Hepatitis.B,
           'Measles Cases per 1000' = Measles,
           '5 year Mortality per 1000' = under.five.deaths,
           'Percentage Vax Polio' = Polio,
           'Percentage of GDP Spent on Healthcare' = Total.expenditure,
           'Percentage Vax DTP3' = Diphtheria,
           'Infant HIV Deaths per 1000' = HIV.AIDS,
           'GDP per Person' = GDP,
           'Average Education (Years)' = Schooling,
           'Percentage Below Ideal BMI' = Thinness) %>%
    pivot_longer(., everything(), names_to = 'Demographic', values_to = 'Value')
    

```















